# Module 3: Generalized Linear Models

## Lecture

<iframe width="640" height="360" src="https://www.youtube.com/embed/KQZ3LXzlKLc?si=Z0inqbuG6OSXQfiu" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>

<br>

![Module 3: Generalized Linear Models](content-files/module-3-lecture.pdf){width=100% height=900}\

## Lab

In this module we're going to cover:

* Reading data from files and evaluating missingness
* Creating publication-quality plots with `ggplot2`
* Fitting models with binary outcomes using generalized linear models
 
### Essential R: Reading Tables From Files, Merging, Basic Data Exploration

For this we're going to use two data files available in the course data directory.

Download the following datasets and put them in your R working directory.

* `tomerge1.csv` : comma-separated values
* `tomerge2.txt` : space-delimited values

Use `read.delim()` to read in tables. Note the use of the `sep=` parameter to indicate what the column separator is:
```{r, message=FALSE}
x1 <- read.delim("datasets/tomerge1.csv",sep=",")
head(x1)
```

```{r, message=FALSE}
x2 <- read.delim("datasets/tomerge2.txt",sep=" ")
head(x2)
```

Use `merge()` to combine the two tables by sample ID. Note the use of `by.x` and `by.y` to tell `merge()` which columns are equivalent:
```{r, class.source="codeblock", eval=TRUE}
x_merge <- merge(x1, x2, by.x = "Sample_ID", by.y = "sampleID")
head(x_merge)
```

We're going to use a popular dataset for data analysis, pertaining to survival of passengers aboard the Titanic.

Download the following dataset and put them in your R working directory.

* `titanic.csv`

Let's read in the data from a file:
```{r, message=FALSE}
dat <- read.delim(
    "datasets/titanic.csv",
    sep="," # indicate the column separator
    ) 
```

Examine the columns:
```{r, message=FALSE}
head(dat)
```

Some of the columns are categorical, use `table()` to look at the tallies. Examine the columns:
```{r, message=FALSE}
table(dat$Survived)
table(dat$Pclass)
```

Use `summary()` to look at continuous-valued data:
```{r, message=FALSE}
summary(dat$Age)
```

Notice that there are 177 `NA` (missing) values for age. Let's visualize the missing data more systematically.

### Explore Missing Data

For this let's use a little script that converts a table into black and white squares to visualize missing data. For this, install the `plotrix` package.
```{r, message=FALSE}
if (!requireNamespace("plotrix", quietly = TRUE)) install.packages("plotrix")

suppressMessages(library(plotrix))

#' show data missingness as a chequered matrix
#' 
#' @param x (matrix) data matrix.
#' @param outFile (char) path to file for printing graph
#' @param wd (numeric) width in inches
#' @param ht (numeric) height in inches
#' @return plots missingness matrix to file
#' @import plotrix
#' @export
plotMissMat <- function(x,xlab="columns",
		ylab="rows",border=NA) {
	
	x <- !is.na(x)
	class(x) <- "numeric"
	color2D.matplot(x,show.values=FALSE,axes=FALSE,
		cs1=c(1,0),cs2=c(1,0),cs3=c(1,0),border=border,
		cex=0.8,
		xlab=xlab,ylab=ylab)
}
```

Let's look at the missingness in the Titanic dataset. Missing data is shown as a white cell, and non-missing data is shown in black.
```{r, message=FALSE}
plotMissMat(dat)
```

We can see a column with many missing values. This is probably the "age" data. 
Let's count the number of missing values on a per-column level.

For this we combine `is.na()`, which returns a TRUE/FALSE value for NA values, and `colSums()` which adds up the TRUE values down the columns.
```{r count-missing, message=FALSE}
colSums(is.na(dat))
```

This confirms that `Age` is the only column with missing data.

Now let's explore the data using plots.

### Essential R: Plots with `ggplot2`

`ggplot2` is a popular plotting package that uses an additive paradigm to build plots. 
 
Useful websites:
* [ggplot2 cheatsheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html)
* The [ggplot2 website](https://ggplot2.tidyverse.org/) is a wealth of reference information, so here we just touch on the basics to get you started.

Anytime you need to generate a specific kind of plot, the website will most likely have documentation for how to achieve it.

Let's start by creating a scatterplot with two **continuous variables**. For this let's load a dataset measuring statistics around quality of life in US states in the late 70's:
```{r, message=FALSE}
state.x77 <- as.data.frame(state.x77)
head(state.x77)
```

Create a base plot using the `ggplot` function. Then "add" a scatterplot to it. Notice that the plot has been assigned to a variable named `p`. 

This setup is standard for `ggplot2` and allows multiple visualizations to be applied to the same base plot.

We use `aes` to tell ggplot what the `x` and `y` axes are, and later, if we want to colour-code by a particular column.
```{r load-states-data, message=FALSE}
library(ggplot2)
p <- ggplot(state.x77, 
    aes(x = Illiteracy,y = Income)
    )

p <- p + geom_point() # scatter plot
p
```

Now let's add confidence intervals:
```{r states-geomsmooth}
p + geom_smooth()
```

It looks like there is a negative relationship between illiteracy and income. We can confirm this by looking at correlation:
```{r test-cor, message=FALSE}
x <- state.x77$Illiteracy
y <- state.x77$Income
cor.test(x,y)
cor.test(x,y)$p.value
```

Let's now examine **categorical variables**. In the titanic set, let's look at the fare paid based on passenger class:
```{r boxplot, message=FALSE}
p <- ggplot(dat)
p + geom_boxplot(
    aes(x = factor(Pclass), # "factor()" makes a data column a categorical variable
        y = Fare))
```

We can use barplots to examine counts and proportions. Let's look at number of survivors, split by passenger class.

Here the `fill` command is used to to split each barplot by the category, "Survived".
So you can see number of passengers by "Pclass" split by survival.
```{r  barplot, message=FALSE}
p + geom_bar(
    aes(fill=factor(Survived), 
        x = Pclass)
)
```

The plot above shows count data. Let's convert this to proportions. We can see that the fraction of non-survivors in "Class 3" is high.
```{r barplot2 , message=FALSE}
p + geom_bar(
    aes(fill=factor(Survived), x = Pclass),
    position = "fill"
)
```

How about males versus females?
```{r finish-titanic, message=FALSE}
p + geom_bar(
    aes(fill=factor(Survived), x = Sex),
    position = "fill"
)
```

**Exercise for later:** Try other `ggplot` functions on these data or on a small data table of your project.
(avoid using large genomics datasets, because those are going to be harder to interpret)

### Fit Binary Response Variable Using `glm()` and Logistic Regression

Let's fit a model to a binary outcome. For this we load a dataset that measures physiological variables in a cohort of Pima Indians.
```{r load-pima, message=FALSE}
library(mlbench)
data(PimaIndiansDiabetes2)
# type ?PimaIndiansDiabetes2 to learn more about the dataset.

dat <- PimaIndiansDiabetes2
head(dat)
```

Let's look at the impact of blood glucose levels on diabetes diagnosis. 

First let's make a scatterplot. Could there be a relationship?
```{r plot-diabetes-vs-glucose, message=FALSE}
p <- ggplot(dat, aes(x = glucose, y = factor(diabetes)))
p + geom_point()
```

Could this be fit with a linear model? 

Is there a continuous line that could reasonably fit the data?
```{r, message=FALSE}
p <- ggplot(dat, aes(x = glucose, y = factor(diabetes)))
p + geom_point() + geom_smooth()
```

This is a situation where a logistic regression model would be an appropriate choice because of the binary outcome. 

We're going to use `glm()` to fit a model to these data:
```{r diabetes-model , message=FALSE}
mod <- glm(factor(diabetes)~glucose, 
    dat,
    family = "binomial" # set to model binary outcome
)
summary(mod)
```

Which factors explain a diabetes diagnosis?

What if we include a couple other factors?
```{r diabetes-model-expanded, message=FALSE}
mod <- glm(factor(diabetes)~ glucose + pregnant + age + pressure + triceps,
    dat,
    family = "binomial")
summary(mod)
```

```{block,type="rmd-caution"}
Note: This morning's session only intends to introduce you to fitting non-linear models. 

In practice you may need to do more work to test multiple models to ascertain best fits your data, using measures such as goodness-of-fit. You will also likely compute odds ratio (odds of increased Y per unit increase X), which is out of scope for the current tutorial. 

We strongly recommend that you learn these topics before applying these methods to your own data.
```

### Bonus Exercise

Now let’s apply the ideas above to a dataset for classifying a breast cell as being either benign or malignant.

```{r, message=FALSE}
data(BreastCancer)
bc <- BreastCancer
for (k in 2:10) # altered for current lab
    bc[,k] <- as.numeric(bc[,k]) 
head(bc)
```

Learn more about the dataset:
```{r, message=FALSE}
?BreastCancer
```

For your exercise, answer the following questions:

* Is there missing data?
* Which columns are
* Use plots to explore the relationship between explanatory variables.
* Fit a logistic model to identify which factors explain class (benign vs. malignant).

### Bonus Exercise Results

```{r, message=FALSE}
data(BreastCancer)
bc <- BreastCancer
for (k in 2:10) # altered for current lab
    bc[,k] <- as.numeric(bc[,k]) 
head(bc)
```

Explore missingness:
```{r}
suppressMessages(require(plotrix))

#' show data missingness as a chequered matrix
#' 
#' @param x (matrix) data matrix.
#' @param outFile (char) path to file for printing graph
#' @param wd (numeric) width in inches
#' @param ht (numeric) height in inches
#' @return plots missingness matrix to file
#' @import plotrix
#' @export
plotMissMat <- function(x,xlab="columns",
        ylab="rows",border=NA) {
    
    x <- !is.na(x)
    class(x) <- "numeric"
    color2D.matplot(x,show.values=FALSE,axes=FALSE,
        cs1=c(1,0),cs2=c(1,0),cs3=c(1,0),border=border,
        cex=0.8,
        xlab=xlab,ylab=ylab)
}
```

Explore missingness:
```{r, message=FALSE}
plotMissMat(bc)
```

```{r, message=FALSE}
colSums(is.na(bc))
```

Plot relationship between variables:
```{r, message=FALSE}
require(ggplot2)

ggplot(bc,aes(Cell.size,Normal.nucleoli)) + geom_point()
```

```{r, message=FALSE}
ggplot(bc) + geom_boxplot(aes(factor(Class), Normal.nucleoli))
```

```{r, message=FALSE}
ggplot(bc) + geom_boxplot(aes(factor(Class), Cl.thickness))
```

```{r, message=FALSE}
ggplot(bc) + geom_boxplot(aes(factor(Class), Bare.nuclei))
```

Fit a binary outcome model:
```{r, message=FALSE}
mod <- glm(
    Class ~ Cl.thickness + Bare.nuclei + Normal.nucleoli + Mitoses + Bl.cromatin, 
    bc,
    family="binomial")
summary(mod)
```
